SPIDERMAN: an open source code to model phase curves and secondary eclipses.

### 2017-07-19 ms.tex
overleaf

Using a geometrical algorithm, the code solves exactly the area of sections of the disc of the planet that are occulted by the star. The code is written in C and optimised to run quickly, with no loss in precision. Approximately 1000 models can be generated per second with typical parameters, making Markov Chain Monte Carlo analyses practiable.

Using an analytical approximation to the results of General Circulation Models, we apply the code to archival data on the phase curve of WASP-43b. The model provides a good fit to the data, however, the best fitting model significantly overpredicts the temperature of the night-side. We speculate that this could be due to the presence of clouds on the night side of the planet.

We test statistically for variation of the map parameters as a function of wavelength and find no statistically strong confirmation of a dependence.

@@

displayed visually in figure \ref{fig:harmonics}

@@

The simplest physical model that one construct.

@@

\textsc{spiderman} is an open source project and is being actively developed on github. The code is available on PyPi, and the latest stable version can easily by installed by simply running the command

\begin{Verbatim}[frame=single]
> pip install spiderman-package
\end{Verbatim}

Or, to access the development version from github:

\begin{Verbatim}[frame=single]
> git clone https://github.com/tomlouden/SPIDERMAN
> cd SPIDERMAN
> sudo python setup.py install
\end{Verbatim}

Full installation instructions and other information are available in the documentation, at \url{http://spiderman.readthedocs.io}

A model instance is initialised by typing

\begin{Verbatim}[frame=single]
> spider_params = spiderman.ModelParams(MODEL)
\end{Verbatim}

Where MODEL is the chosen surface brightness distribution. The model parameters are then entered, and from there, generating a lightcurve to compare to data is as simple as entering

\begin{Verbatim}[frame=single]
> spider_params.lightcurve(t)
\end{Verbatim}

Where t an array of timepoints where the model is to be evaluated.

\textsc{spiderman} comes prepackaged with several plotting scripts to make visualisation of the model simple.

Typing the command

\begin{Verbatim}[frame=single]
> spider_params.plot_planet(t)
\end{Verbatim}

Will display the planet from the perspective of the observer on Earth at the desired phase or Julian date (t). If the chosen model uses a temperature distribution, then these plots can be made for both the temperature and the calculated flux distribution.

typing the command

\begin{Verbatim}[frame=single]
> spider_params.plot_system(t)
\end{Verbatim}

will generate a plot like figure \ref{fig:plot_system}, which is a scale diagram of the system. The desired phase or julian date can be given as an argument. These plots can be very useful for visualising the physical impacts of the chosen parameters.

@@

\textsc{spiderman} has been designed to ensure that it is easy to visualise the appearance of the model at all stages.

\subsection{Performance}\label{sec:performance}

@@

As a test of the performance of the code the model was applied to a phase curve and secondary eclipse of WASP-43b taken with HST WFC3. These data had previously been published in \citet{Stevenson2014}. However, in this paper the phase curve and the secondary eclipse are treated as separate phenomena, when of course, they are both manifestations of the same underlying brightness distribution. Indeed, the bright dayside of the planet is the main driver of the phase curve, and is also the part occulted by the stellar disc. It is therefore more parsimonious, if at all possible, to use a model that accounts for both of these observations simultaneously. This also reduces the total number of parameters needed, which implies a better information content.

@@

Fortunately, the phasing of the observations is such that the three visits cover very different ranges of the phase curve, which serves to alleviate the degeneracy.

@@

The results of both runs are presented In table \ref{tab:results}. There is very little difference between the fitted parameters with and without limb darkening. The value of $T_n$ is lower by more than $3\sigma$ in the fit with planetary limb darkening, but the other parameters in the physical model, $\Delta T$ and $\xi$ are consistent to within $2\sigma$ between the two runs. it is particularly worth noting how apparently unaffected $\xi$ is by the presence of limb darkening, the difference being much less than $1\sigma$. This makes sense, as $\xi$ effectively parametrises the asymmetry in the flux distribution, while the limb darkening is symmetrical. This is fortunate, as $\xi$ is the most physically interesting of the model parameter, and is directly tied to the details of heat transport in the atmosphere. The insensitivity of this parameter to limb darkening is important, as it shows that it can be recovered reliably even in ignorance of the vertical temperature structure of the planet's atmosphere.

@@

It is also interesting to check the correlation of this $\xi$ parameter with the other model parameters, this can be seen in the corner plot in figure \ref{fig:triangle corner sub}. $T_s$ and $\Delta T$ are clearly correlated parameters, which is logical, as between them they determine the average temperature of the dayside, which controls the amplitude of the phase curve. Crucially, $\xi$ is not strongly degenerate with the other parameters. This indicates that the $\xi$ parameter is robust to any errors in the absolute flux level of the system, due to, for example, not modelling the instrument response curve.

@@

An intriguing difference between our model fit and the previous fit by \citet{Stevenson2014} is that we are able to infer a nightside temperature, where they were only able to place upper limits. To demonstrate that the flux from the nightside is a real feature in the model fit, and not an abstract parameterisation of some other feature, we investigate the flux level just outside of the primary transit. If the nightside has indeed been significantly detected, the flux here should be significantly different to the star-only flux visible during the secondary eclipse (which is 1.0 by definition). Drawing 10,000 lightcurves from the parameters in the non-limb darkened MCMC posterior, we plot the median, 1 sigma and 3 sigma contours for the lightcurve in figure \ref{fig:model}. The flux both just before, and just after the primary transit is significantly greater than 1.0 at the 3 sigma level, indicating that there is indeed flux detected from the night side of the planet.

@@

As a test of the physical nature of this nightside flux, we repeat the fit with an additional model parameter which allows \emph{planetary} limb darkening or brightening.

Repeating this test with the limb darkened fit does not find significant night side flux, which casts doubt on the significance of this detection.

inhibit limb brightening or limb darkening \citep{Stevenson2014}. However, despite this unexpected result, as previously discussed, there is no significant difference to the physical model fit of \citep{Zhang2016}, which raises our confidence in the robustness of the fit to any potential errors in the symmetric radial component of the surface map.

@@

@@

We demonstrate the use of \textsc{spiderman} by applying the code to the case of WASP-43b. We use a physically motivated analytical model presented by \citet{zhang2016} which captures the major features of exoplanet phase curves. We find that the model provides a good fit to the data, but overpredicts the nightside flux compared to non-informative models. We hypothesise that this may be due to the presence of clouds on the planet's nightside.

@@

%\begin{figure*}
%\begin{center}
%\includegraphics[width=\columnwidth]{img/free_parameterstemp_map.pdf}
%\includegraphics[width=\columnwidth]{img/free_parameterstemp_errs.pdf}
%\caption{The underlying temperature distributions generated by the zhang model.}
%\label{fig:best_fit_temp}
%\end{center}
%\end{figure*}

>>>>

Using a geometrical algorithm, the code solves exactly the area of sections of the disc of the planet that are occulted by the star. The code is written in C and optimised to run quickly, with no loss in precision. Approximately 1000 models can be generated per second with typical parameters, making Markov Chain Monte Carlo analyses practicable.

Using an analytical approximation to the results of General Circulation Models, we apply the code to archival data on the phase curve of WASP-43b. The model provides a good fit to the data, however, the best fitting model overpredicts the temperature of the night-side. We speculate that this could be due to the presence of clouds on the night side of the planet.

We test statistically for variation of the map parameters as a function of wavelength and find no statistically strong correlations.

@@

A completely non-physics dependent model is a sum of spherical harmonics. This has been used before by ((cite)) an example generated by \textsc{spiderman} is displayed in figure \ref{fig:harmonics}

@@

The simplest physically model that one construct involve defining a bulk planetary temperature or flux, and then a "Hotspot" where the contrast and offset from the substellar point in longitude and latitude are model parameters. Models where the day and the night side have different fluxes are a subset of these models.

@@

\textsc{spiderman} is an open source project and is being actively developed on github. The code is available on PyPi and github at \url{https://github.com/tomlouden/spiderman}.

%\begin{Verbatim}[frame=single]
%> pip install spiderman-package
%\end{Verbatim}

%Or, to access the development version from github:

%\begin{Verbatim}[frame=single]
%> git clone https://github.com/tomlouden/SPIDERMAN
%> cd SPIDERMAN
%> sudo python setup.py install
%\end{Verbatim}

Installation instructions and documentation are available at \url{http://spiderman.readthedocs.io}

%A model instance is initialised by typing

%\begin{Verbatim}[frame=single]
%> spider_params = spiderman.ModelParams(MODEL)
%\end{Verbatim}

%Where MODEL is the chosen surface brightness distribution. The model parameters are then entered, and from there, generating a lightcurve to compare to data is as simple as entering

%\begin{Verbatim}[frame=single]
%> spider_params.lightcurve(t)
%\end{Verbatim}

%Where t an array of timepoints where the model is to be evaluated.

%\textsc{spiderman} comes prepackaged with several plotting scripts to make visualisation of the model simple.

%Typing the command

%\begin{Verbatim}[frame=single]
%> spider_params.plot_planet(t)
%\end{Verbatim}

%\begin{Verbatim}[frame=single]
%> spider_params.plot_system(t)
%\end{Verbatim}

As well as lightcurve generation the code includes plotting functions to visualise the brightness or temperature distributions. An example plot is given in \ref{fig:plot_system}, which is a scale diagram of the system at a specified time or phase. Plots like this are useful for visualising the effects of varying the parameters.

@@

@@

As a test of the performance of the code the model was applied to a phase curve and secondary eclipse of WASP-43b taken with HST WFC3. These data have previously been published in \citet{Stevenson2014}. However, in this paper the phase curve and the secondary eclipse are treated as separate phenomena, when of course, they are both manifestations of the same underlying brightness distribution. Indeed, the bright dayside of the planet is the main driver of the phase curve, and is also the part occulted by the stellar disc. It is therefore more parsimonious, if at all possible, to use a model that accounts for both of these observations simultaneously. This also reduces the total number of parameters needed, which implies a better information content.

@@

Fortunately, the phasing of the observations is such that the three visits cover different portions of the phase curve, which helps to alleviate the degeneracy.

@@

The results of both runs are presented In table \ref{tab:results}. With the exception of $T_n$, which is lower by more than $3\sigma$ in the fit with planetary limb darkening, the physical parameters are consistent. In particular, $\xi$ seems only weakly affected by the presence of limb darkening, the difference being much less than $1\sigma$. This is not unexpected, as $\xi$ effectively parametrises the asymmetry in the flux distribution, while the limb darkening is symmetrical. $\xi$ is directly tied to the details of heat transport in the atmosphere, so the insensitivity of this parameter to limb darkening is useful, as it shows that it can be recovered even with systematic uncertainties in the flux distribution, so long as they are symmetric.

@@

The correlation between the physical model with the other model parameters can be seen in figure \ref{fig:triangle corner sub}. $T_s$ and $\Delta T$ are clearly correlated parameters, since between them they determine the average temperature of the dayside and hence the amplitude of the phase curve. $\xi$ is not strongly degenerate with the other parameters. This indicates that the $\xi$ parameter is robust to any errors in the absolute flux level of the system.

@@

The model fit implies that the night-side of the planet is significantly hotter than expected at 1290 K \citet{Stevenson2014}. If the nightside has been significantly detected, the flux immediately before and after primary transit should be significantly different to the star-only flux visible during the secondary eclipse. Drawing 10,000 lightcurves from the parameters in the non-limb darkened MCMC posterior, we plot the median, 1 sigma and 3 sigma contours for the lightcurve in figure \ref{fig:model}. The flux is found to be significantly greater than 1.0 at the 3 sigma level.

@@

((Laura - comment on how bright nightside would be inconsistent with eclipse spectra?))

Repeating this test with the limb darkened fit finds a significantly fainter night side flux, suggesting that the nightside flux is an artifact of using a simple physical model. This may be due to a correlation between model parameters making the fit inflexible and the simultaneous fit with the systematics model. Alternatively, the model may be unable to parametrise a feature of the nightside that is blocking the emitted flux, such as clouds.

However, as previously pointed out, the $\xi$ parameter does not seem to be effected, which raises our confidence in the robustness of the fit to any potential errors in the symmetric radial component of the surface map.

@@

\subsubsection{Spectrally resolved}

@@

We demonstrate the use of \textsc{spiderman} by applying the code to the case of WASP-43b. We use a physically motivated analytical model presented by \citet{Zhang2016} which captures the major features of exoplanet phase curves. We find that the model provides a good fit to the data, but overpredicts the nightside flux compared to non-informative models. We hypothesise that this may be due to the presence of clouds on the planet's nightside.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=\columnwidth]{img/free_parameterstemp_map.pdf}
\includegraphics[width=\columnwidth]{img/free_parameterstemp_errs.pdf}
\caption{The underlying temperature distributions generated by the zhang model.}
\label{fig:best_fit_temp}
\end{center}
\end{figure*}
